Solving Collective Spin Hamiltonians with Decoherence

### 2018-08-19 soc_squeezing/cumulant_expansions.tex
more work on the same commutator

\newcommand{\D}{\mathcal{D}}
\renewcommand{\O}{\mathcal{O}}

@@

\section{Spin operator identities}
\label{sec:identities}

This text makes ubiquitous use of various spin operator identities; we
collect and derive particularly useful identities here for reference.
The basic commutation relations between single-weight spin operators
are
  \f12 \sp{\sigma_j^z,\sigma_k^\mu}_-
  &= \delta_{jk} \mu \sigma_j^\mu,
  &
  \sp{S_z,\sigma_j^\mu}_-
  &= \f12 \sp{\sigma_j^z,S_\mu}_- = \mu \sigma_j^\mu,
  &
  \sp{S_z,S_\mu}_-
  &= \mu S_\mu,
  \label{eq:comm_z_mu} \\
  \sp{\sigma_j^\mu,\sigma_k^\nu}_-
  &= \delta_{jk} \delta_{\mu,-\nu} \mu \sigma_j^z,
  &
  \sp{S_\mu,\sigma_j^\nu}_-
  &= \sp{\sigma_j^\mu,S_\nu}_-
  = \delta_{\mu,-\nu} \mu \sigma_j^z,
  &
  \sp{S_\mu,S_\nu}_-
  &= \delta_{\mu,-\nu} 2 \mu S_z.
  \label{eq:comm_mu_nu}
These relations can be used to inductively determine commutation
relations between higher-weight spin operators.  In particular, we can
find that
  S_z S_\mu^m
  = \p{\mu S_\mu + S_\mu S_z} S_\mu^{m-1}
  = \mu S_\mu^m + S_\mu S_z S_\mu^{m-1}
  = 2 \mu S_\mu^m + S_\mu S_z S_\mu^{m-2}
  = S_\mu^m \p{\mu m + S_z},
  \label{eq:push_z_mu}
and in turn
  S_z S_\mu^m S_\nu^n
  = S_\mu^m \p{\mu m + S_z} S_\nu^n
  = S_\mu^m S_\nu^n \p{\mu m + \nu n + S_z}.
  \label{eq:push_z_mu_nu}
Repeating this process $\ell$ times tells us
  S_z^\ell S_\mu^m S_\nu^n
  = S_\mu^m S_\nu^n \p{\mu m + \nu n + S_z}^\ell,
  \sp{S_z^\ell, S_\mu^m S_\nu^n}_-
  = S_\mu^m S_\nu^n \p{\sp{\mu m + \nu n + S_z}^\ell - S_z^\ell}
  = S_\mu^m S_\nu^n  \sum_{k=0}^{\ell-1} {\ell \choose k}
  \p{\mu m + \nu n}^{\ell-k} S_z^k.
  \label{eq:comm_z_mu_nu_pow}

@@

transverse to the $z$ axis, we compute (with $\bmu\equiv-\mu$ for
brevity)

@@

  + S_\bmu^n S_\mu,

@@

where

@@

  2 \sum_{k=0}^{n-1} S_\bmu^k S_z S_\bmu^{n-k-1}
  = 2 \sum_{k=0}^{n-1} S_\bmu^k S_\bmu^{n-k-1} \p{\bmu\sp{n-k-1} + S_z}
  = n S_\bmu^{n-1}\p{\bmu\sp{n-1} + 2S_z},

@@

  = n S_\bmu^{n-1}\p{\bmu\sp{n-1} + 2S_z}.

@@

Finally, we attempt to compute

@@

  &= S_\mu^{m-1} \sp{\mu n S_\bmu^{n-1}\p{\bmu\sp{n-1} + 2S_z}

@@

  &= \mu n \sp{S_\mu^{m-1} S_\bmu^{n-1} \p{\bmu\sp{n-1} + 2S_z}}

@@

  &= \mu n \sum_{\ell=0}^{m-1} S_\mu^{m-\ell-1} S_\bmu^{n-1}
  \p{\bmu\sp{n-1} + 2S_z} S_\mu^\ell + S_\bmu^n S_\mu^m \\
  &= C_{\mu mn} + S_\bmu^n S_\mu^m,

@@

where

@@

  &\equiv \mu n \sum_{\ell=0}^{m-1}
  S_\mu^{m-\ell-1} S_\bmu^{n-1} S_\mu^\ell
  \p{\bmu\sp{n-1} + 2\mu\ell + 2S_z} \\
  S_\mu^{m-\ell-1} S_\bmu^{n-1} S_\mu^\ell

@@

  + n \sum_{\ell=0}^{m-1} C_{\mu,m-\ell-1,n-1}

@@

  \label{eq:C_mn_mu}

@@

@@

  \label{eq:D_mn_mu}

@@

We generally want all spin operators to be in a standard order, so we
now wish to push all $S_z$ operators in \eqref{eq:C_mn_mu} to the
right.  To do so, we define

@@

where we can write

@@

  &= \p{m-\ell}n S_\bmu^{n-1} S_\mu^{m-1} \p{\ell+m-n+2\mu S_z}.

@@

We now rearrange

@@

  &= D_{\mu,m-\ell,n} S_\mu^\ell + n \sum_{k=0}^{m-\ell-1}

@@

  &= D_{\mu mn}^{(\ell)} + n \sum_{k=0}^{m-\ell-1}
  C_{\mu,m-\ell-k-1,n-1} S_\mu^{k+\ell} \p{2\ell+2k+1-n+2\mu S_z} \\
  &= D_{\mu mn}^{(\ell)} + n \sum_{k=0}^{m-\ell-1}
  C_{\mu,m-1,n-1}^{(\ell+k)} \p{2\sp{\ell+k}+1-n+2\mu S_z} \\
  &= D_{\mu mn}^{(\ell)} + n \sum_{k=\ell}^{m-1}
  C_{\mu,m-1,n-1}^{(k)} \p{2k+1-n+2\mu S_z},

@@

which provides a recursive formula for
$\sp{S_\mu^m,S_\bmu^{n}}_- = C_{\mu mn} = C_{\mu mn}^{(0)}$ in which
all operators are in a standard order.  While solving this formula may
be more trouble than it's worth right now, to derive a system of
equations for the two-axis twisting Hamiltonian we need to know
$\sp{S_\mu^2,S_\bmu^n}_-$, so we expand

@@

  \sp{S_\mu^2,S_\bmu^n}_-
  = D_{\mu,2,n} + n D_{\mu,1,n-1} \p{1 - n + 2\mu S_z}

@@

\begin{align}
  C_{\mu mn}^{(\ell)}
  = D_{\mu mn}^{(\ell)} + n \sum_{k=\ell}^{m-1}
  C_{\mu,m-1,n-1}^{(k)} \p{2k+1-n+2\mu S_z}
  = D_{\mu mn}^{(\ell)} + \sum_{p=1}^{\min\p{m,n}}
  \mathcal{E}_{\mu mn}^{(\ell,p)}
\end{align}
\begin{align}
  \mathcal{E}_{\mu mn}^{(\ell,1)}
  = n \sum_{k_1=\ell}^{m-1} D_{\mu,m-1,n-1}^{(k_1)}
  \p{2k_1-\sp{n-1}+2\mu S_z}
\end{align}
\begin{align}
  \mathcal{E}_{\mu mn}^{(\ell,2)}
  = n\p{n-1} \sum_{k_1=\ell}^{m-1} \sum_{k_2=k_1}^{m-1}
  D_{\mu,m-2,n-2}^{(k_2)}
  \p{2k_2-\sp{n-2}+2\mu S_z} \p{2k_1-\sp{n-1}+2\mu S_z}
\end{align}
\begin{align}
  \mathcal{E}_{\mu mn}^{(k_0,p)}
  = 2^p n \cdots \p{n-p+1}
  \sum_{k_1=k_0}^{m-1} \cdots \sum_{k_p=k_{p-1}}^{m-1}
  D_{\mu,m-p,n-p}^{(k_p)}
  \sp{f\p{k_1,1}+\mu S_z} \cdots \sp{f\p{k_p,p}+\mu S_z}
\end{align}
where
\begin{align}
  f\p{k,p} = k - \f12\p{n-p}.
\end{align}

>>>>

\newcommand{\C}{\mathcal{C}}
\newcommand{\D}{\mathcal{D}}
\renewcommand{\O}{\mathcal{O}}

@@

\section{Basic spin operator identities}
\label{sec:identities}

This text makes ubiquitous use of various spin operator identities; we
collect and derive particularly useful identities here for reference.
The elementary commutation relations between single-weight spin
operators are
\begin{align}
  \f12 \sp{\sigma_j^z,\sigma_k^\mu}_-
  &= \delta_{jk} \mu \sigma_j^\mu,
  &
  \sp{S_z,\sigma_j^\mu}_-
  &= \f12 \sp{\sigma_j^z,S_\mu}_- = \mu \sigma_j^\mu,
  &
  \sp{S_z,S_\mu}_-
  &= \mu S_\mu,
  \label{eq:comm_z_mu} \\
  \sp{\sigma_j^\mu,\sigma_k^\nu}_-
  &= \delta_{jk} \delta_{\mu,-\nu} \mu \sigma_j^z,
  &
  \sp{S_\mu,\sigma_j^\nu}_-
  &= \sp{\sigma_j^\mu,S_\nu}_-
  = \delta_{\mu,-\nu} \mu \sigma_j^z,
  &
  \sp{S_\mu,S_\nu}_-
  &= \delta_{\mu,-\nu} 2 \mu S_z.
  \label{eq:comm_mu_nu}
\end{align}
These relations can be used to inductively determine commutation
relations between higher-weight spin operators.  In particular, we can
find that
\begin{align}
  S_z S_\mu^m
  = \p{\mu S_\mu + S_\mu S_z} S_\mu^{m-1}
  = \mu S_\mu^m + S_\mu S_z S_\mu^{m-1}
  = 2 \mu S_\mu^m + S_\mu S_z S_\mu^{m-2}
  = S_\mu^m \p{\mu m + S_z},
  \label{eq:push_z_mu}
\end{align}
and in turn
\begin{align}
  S_z S_\mu^m S_\nu^n
  = S_\mu^m \p{\mu m + S_z} S_\nu^n
  = S_\mu^m S_\nu^n \p{\mu m + \nu n + S_z}.
  \label{eq:push_z_mu_nu}
\end{align}
Repeating this process $\ell$ times tells us
\begin{align}
  S_z^\ell S_\mu^m S_\nu^n
  = S_\mu^m S_\nu^n \p{\mu m + \nu n + S_z}^\ell,
\end{align}
which implies
\begin{align}
  \sp{S_z^\ell, S_\mu^m S_\nu^n}_-
  = S_\mu^m S_\nu^n \p{\sp{\mu m + \nu n + S_z}^\ell - S_z^\ell}
  = S_\mu^m S_\nu^n  \sum_{k=0}^{\ell-1} {\ell \choose k}
  \p{\mu m + \nu n}^{\ell-k} S_z^k.
  \label{eq:comm_z_mu_nu_pow}
\end{align}

\section{Commutators between high-weight transverse spin operators}

@@

transverse to the decoherence (i.e.~$z$) axis, we compute (with
$\bmu\equiv-\mu$ for brevity)

@@

  + S_\bmu^n S_\mu.
  \label{eq:comm_mu_nu_single}

@@

While \eqref{eq:comm_mu_nu_single} gives us the commutator
$\sp{S_\mu,S_\bmu^n}_-$, we would like to enforce an ordering on
products of collective spin operators, which allows us to only keep
track of high-weight operators which are linearly independent.  We
(arbitrarily) choose to impose an ordering with all $S_z$ operators on
the right.  This choice of ordering compels us to expand

@@

  2 \mu \sum_{k=0}^{n-1} S_\bmu^k S_z S_\bmu^{n-k-1}
  = 2 \mu \sum_{k=0}^{n-1}
  S_\bmu^k S_\bmu^{n-k-1} \p{\bmu\sp{n-k-1} + S_z}
  = n S_\bmu^{n-1}\p{1 - n + 2\mu S_z},

@@

  = n S_\bmu^{n-1}\p{1 - n + 2\mu S_z}.

@@

As the next logical step, we take on the task of computing

@@

  &= S_\mu^{m-1} \sp{n S_\bmu^{n-1}\p{1 - n + 2\mu S_z}

@@

  &= n \sp{S_\mu^{m-1} S_\bmu^{n-1} \p{1 - n + 2\mu S_z}}

@@

  &= n \sum_{\ell=0}^{m-1} S_\mu^{m-\ell-1} S_\bmu^{n-1}
  \p{1 - n + 2\mu S_z} S_\mu^\ell + S_\bmu^n S_\mu^m,

@@

  \sp{S_\mu^m, S_\bmu^n}_-
  \equiv n \sum_{\ell=0}^{m-1} S_\mu^{m-\ell-1} S_\bmu^{n-1}
  \p{1 - n + 2\mu S_z} S_\mu^\ell.
We now need rearrange the operators in $C_{\mu mn}$ into a standard
order, which means pushing all $S_z$ operators to the right and, for
the purposes of this calculation, all $S_\bmu$ operators on the left.
We begin by pushing $S_\mu^\ell$ to the left of $S_z$, which takes
$2\mu S_z\to 2\mu S_z+2\ell$, and then push $S_\mu^{m-\ell-1}$ to the
right of $S_\bmu^{n-1}$, giving us

@@

@@

  + n \sum_{\ell=0}^{m-2} C_{\mu,m-\ell-1,n-1}

@@

  \label{eq:C_mn}

@@

where we have dropped the last term in the final sum because
$C_{\mu,m-\ell-1,n-1}=\sp{S_\mu^{m-\ell-1},S_\bmu^{n-1}}_-=0$ if
$\ell=m-1$, and

@@

  \label{eq:D_mn}

@@

To our despair, we have arrived in \eqref{eq:C_mn} to a {\it
  recursive} expression for $C_{\mu mn}$, and we have not even managed
to order the spin operators, as $C_{\mu,m-\ell-1,n-1}$ contains $S_z$
operators which are to the left of $S_\mu^\ell$.  To sort all spin
operators once and for all, we define

@@

which we can expand as

@@

  &= \p{m-\ell}n S_\bmu^{n-1} S_\mu^{m-1} \p{\ell+m-n+2\mu S_z},
  \label{eq:D_mn_ell}

@@

@@

  &= D_{\mu,m-\ell,n} S_\mu^\ell + n \sum_{k=0}^{m-\ell-2}

@@

  &= D_{\mu mn}^{(\ell)} + n \sum_{k=0}^{m-\ell-2}
  C_{\mu,m-\ell-k-1,n-1} S_\mu^{k+\ell} \p{2k+2\ell+1-n+2\mu S_z} \\
  &= D_{\mu mn}^{(\ell)} + n \sum_{k=0}^{m-\ell-2}
  C_{\mu,m-1,n-1}^{(\ell+k)} \p{2\sp{k+\ell}+1-n+2\mu S_z} \\
  &= D_{\mu mn}^{(\ell)} + n \sum_{k=\ell}^{m-2}
  C_{\mu,m-1,n-1}^{(k)} \p{2k+1-n+2\mu S_z}.
  \label{eq:C_mn_ell}

@@

While the resulting expression strongly resembles that in
\eqref{eq:C_mn}, there is one crucial difference: all spin operators
in \eqref{eq:C_mn_ell} have been ordered.  We can now repeatedly
substitute $C_{\mu mn}^{(\ell)}$ into itself, each time decreasing $m$
and $n$ by 1, until one of $m$ or $n$ reaches zero.  Such repeated
substitution yields the expansion

@@

  = D_{\mu mn}^{(\ell)}
  + \sum_{p=1}^{\min\set{m,n}-1} E_{\mu mn}^{(\ell,p)},
where the first terms in the sum are
  E_{\mu mn}^{(\ell,1)}
  = n \sum_{k=\ell}^{m-2} D_{\mu,m-1,n-1}^{(k)}
  \p{2k+1-n+2\mu S_z},
  E_{\mu mn}^{(\ell,2)}
  = n \sum_{k_1=\ell}^{m-2} \p{n-1} \sum_{k_2=k_1}^{m-3}
  D_{\mu,m-2,n-2}^{(k_2)}
  \p{2k_2+2-n+2\mu S_z} \p{2k_1+1-n+2\mu S_z},
and more generally
  E_{\mu mn}^{(k_0,p)}
  = n^{\underline{p}} \sum_{\p{\v k,k_0,m,p}}
  D_{\mu,m-p,n-p}^{(k_p)} \prod_{j=1}^p \p{2k_j+j-n+2\mu S_z}.
  \label{eq:E_mn_kp}
with
$n^{\underline{p}}\equiv
n\p{n-1}\cdots\p{n-p+1}=\prod_{j=0}^{p-1}\p{n-j}$ and
  \sum_{\p{\v k,k_0,m,p}}
  \equiv \sum_{k_1=k_0}^{m-2} \cdots \sum_{k_p=k_{p-1}}^{m-p-1}.
Using \eqref{eq:D_mn_ell} to substitute $D_{\mu,m-p,n-p}^{(k_p)}$ into
\eqref{eq:E_mn_kp} gives us
  E_{\mu mn}^{(k_0,p)}
  &= n^{\underline{p}} S_\bmu^{n-p-1} S_\mu^{m-p-1} \tag*{} \\
  &\qquad \times \sum_{\p{\v k,k_0,m,p}}
  \p{m-p-k_p} \p{n-p} \p{k_p+m-n+2\mu S_z} \tag*{} \\
  &\qquad \times \prod_{j=1}^p \p{2k_j+j-n+2\mu S_z}.
We now wish to pick off the coefficients of each operator in
$E_{\mu mn}^{\p{k_0,p}}$, so we express
  \prod_{j=1}^p \p{2k_j+p-n+2\mu S_z}
  = \sum_{q=0}^p \p{2\mu S_z}^q
  \sum_{c\in\C\p{q,p}} \prod_{\alpha\notin c} \p{2k_\alpha+\alpha-n},
  \label{eq:product_expansion}
where implicitly $\alpha\in\set{1,2,\cdots,p}$, and $\C\p{q,p}$
denotes the set of $q$-combinations (i.e.~subsets with $q$ elements)
of $\set{1,2,\cdots,p}$; the number of elements in $\C\p{q,p}$ is
${p \choose q}$.  Note that for the expansion in
\eqref{eq:product_expansion} to hold, we require that the empty
product
$\prod_{\alpha\notin\set{1,\cdots,p}}\p{2k_\alpha+\alpha-n}=1$.  We
can now write
  E_{\mu mn}^{\p{k_0,p}}
  &= n^{\underline{p}} S_\bmu^{n-p-1} S_\mu^{m-p-1}
  \sum_{q=0}^p \p{2\mu}^q S_z^q \tag*{} \\
  &\qquad \times \sum_{c\in\C\p{q,p}} \sum_{\p{\v k,k_0,m,p}}
  \p{m-p-k_p} \p{n-p} \p{k_p+m-n+2\mu S_z}
  \prod_{\alpha\notin c} \p{2k_\alpha+\alpha-n}.

@@

Solving \eqref{eq:C_mn_ell} may be more trouble than it's worth
right now, but to derive a system of equations for the two-axis
twisting Hamiltonian we need to know $\sp{S_\mu^2,S_\bmu^n}_-$, so we
expand
\begin{align}
  \sp{S_\mu^2,S_\bmu^n}_-
  = C_{\mu mn}
  = D_{\mu,2,n} + n D_{\mu,1,n-1} \p{1 - n + 2\mu S_z}
\end{align}
